---
title: "Data624_Project1_2"
author: "Ritesh Lohiya"
date: "March 28, 2019"
output: html_document
---

```{r}
suppressMessages(suppressWarnings(library(fpp2)))
suppressMessages(suppressWarnings(library(readxl)))
suppressMessages(suppressWarnings(library(seasonal)))
suppressMessages(suppressWarnings(library(rdatamarket)))
suppressMessages(suppressWarnings(library(tseries)))
suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(scales)))
#suppressMessages(suppressWarnings(library(theme_light())))
suppressMessages(suppressWarnings(library(forecast)))
```


##Part B - Forecasting Power, ResidentialCustomerForecastLoad-624.xlsx

####Part B consists of a simple dataset of residential power usage for January 1998 until December 2013.  Your assignment is to model these data and a monthly forecast for 2014.  The data is given in a single file.  The variable 'KWH' is power consumption in Kilowatt hours, the rest is straight forward.    Add this to your existing files above. 

```{r}
# read the data
pu_data <- readxl::read_excel("C:/Users/rites/Documents/GitHub/Data624-Project1/ResidentialCustomerForecastLoad-624.xlsx")

#timeseries
Pu_data_ts =ts(pu_data[,"KWH"],start = c(1998,1),frequency = 12)


```