---
title: "Data624_Project1_2"
author: "Ritesh Lohiya"
date: "March 28, 2019"
output: html_document
---

```{r}
suppressMessages(suppressWarnings(library(fpp2)))
suppressMessages(suppressWarnings(library(readxl)))
suppressMessages(suppressWarnings(library(seasonal)))
suppressMessages(suppressWarnings(library(rdatamarket)))
suppressMessages(suppressWarnings(library(tseries)))
suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(scales)))
#suppressMessages(suppressWarnings(library(theme_light())))
suppressMessages(suppressWarnings(library(forecast)))
```


##Part B - Forecasting Power, ResidentialCustomerForecastLoad-624.xlsx

####Part B consists of a simple dataset of residential power usage for January 1998 until December 2013.  Your assignment is to model these data and a monthly forecast for 2014.  The data is given in a single file.  The variable 'KWH' is power consumption in Kilowatt hours, the rest is straight forward.    Add this to your existing files above. 

```{r}
# read the data
pu_data <- readxl::read_excel("C:/Users/rites/Documents/GitHub/Data624-Project1/ResidentialCustomerForecastLoad-624.xlsx")

#timeseries
pu_data =ts(pu_data[,"KWH"],start = c(1998,1),frequency = 12)
```

#### Data exploration

```{r}
autoplot(pu_data) +
labs(title = "Monthly residential power usage")
```

#### We can see annual seasonality in the data.

```{r}
# Box-cox paramter
pu_lambda <- BoxCox.lambda(pu_data)
pu_trans <- BoxCox(pu_data, pu_lambda)
```

#### Fitting

```{r}
ggtsdisplay(diff(pu_trans, 12), points = FALSE,
main = "Dif lag12 Box-Cox transformed residential power usage")
```

The series is stationary, so no non-seasonal differencing is needed.  The decaying seasonal spikes in the PACF suggests a seasonal AR(1) component, while the very quickly-decaying seasonal spikes in the ACF suggest the possibility of a seasonal MA(1) component.  Spikes in the PACF and ACF at $k = 1$ and $k = 4$ suggest non-seasonal AR(1) or AR(4) components, and non-seasonal MA(1) or MA(4) components.  The function used to select the model with lowest AIC in Part A is redefined for use on the `kWh` timeseries with $D = 1$ and $d = 0$:

```{r}
# redefine function
pu_aic <- function(p, q, P, Q) {
# create model with Box-Cox and specified ARIMA parameters; extract AIC
AIC(Arima(pu_data, order = c(p, 0, q), seasonal = c(P, 1, Q), lambda = pu_lambda))
}
# create possible combinations except all zero & p = q = 4; P = Q = 1 (returns error)
expand.grid(p = c(0, 1, 4), q = c(0, 1, 4), P = 0:1, Q = 0:1) %>%
filter(p > 0 | q > 0 | P > 0 | Q > 0, p < 4 | q < 4 | P < 1 | Q < 1) %>%
# calc AIC for models
mutate(aic = pmap_dbl(list(p, q, P, Q), pu_aic)) %>% 
# return best AIC
slice(which.min(aic))
```


####The minimum AIC value returned is for the ARIMA(1,0,0)(0,1,1) model; this is used:

```{r}
pu_fit <- Arima(pu_data, order = c(1, 0, 0), seasonal = c(0, 1, 1), lambda = pu_lambda)
```

####Ljung-Box test and diagnostic plotting:

```{r}
Box.test(resid(pu_fit), type = "L", fitdf = 3, lag = 12)
ggtsdisplay(resid(pu_fit), points = FALSE,
main = "Residuals for ARIMA(1,0,0)(0,1,1) of residential power usage")
```

####From Ljung-Box test,  p-value > 0.05, but the spikes in ACF & PACF at $k = 3$ and $k = 4$ suggest the possibility of AR(3) or MA(3) components (since the spike at $k = 4$ was addressed above).  Investigation of these does not yield any AIC values lower than that of the above-identified model:

```{r}
expand.grid(p = c(1, 3), q = c(1, 3)) %>%
mutate(aic = pmap_dbl(list(p, q, 0, 1), pu_aic))
```

####The residuals appears to be normally distributed around zero without any significant autocorrelation:

```{r}
ggtsdisplay(resid(pu_fit), points = FALSE, plot.type = "histogram",
main = "Residuals for ARIMA(1,0,0)(0,1,1) of residential power usage")
```

#### Forecasting

Using the ARIMA(1,0,0)(0,1,1) model, the next year (12 months) is forecast, and this forecast is plotted:

```{r}
pu_forecast <- forecast(pu_fit, 12, level = 95)
autoplot(pu_forecast) + 
labs(title = "Forecasted residential enery use",
subtitle = "Using ARIMA(1,0,0)(0,1,1) model", x = "Month", y = "kWh") +
theme(legend.position = "none")
```

#### The forecast shows annual seasonality. 

```{r}
data_frame(`YYYY-MMM` = paste0(2014, "-", month.abb),
PU = pu_forecast$mean) %>% 
write_csv("C:/Users/rites/Documents/GitHub/Data624-Project1/project1_kWh.csv")
```

