---
title: "Data624 Project1_3"
author: "Ritesh Lohiya"
date: "March 29, 2019"
output: html_document
---

```{r}
suppressMessages(suppressWarnings(library(fpp2)))
suppressMessages(suppressWarnings(library(readxl)))
suppressMessages(suppressWarnings(library(seasonal)))
suppressMessages(suppressWarnings(library(rdatamarket)))
suppressMessages(suppressWarnings(library(tseries)))
suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(scales)))
#suppressMessages(suppressWarnings(library(theme_light())))
suppressMessages(suppressWarnings(library(forecast)))
suppressMessages(suppressWarnings(library(lubridate)))
```


##Part C - BONUS, optional (part or all), Waterflow_Pipe1.xlsx and Waterflow_Pipe2.xlsx

####Part C consists of two data sets.  These are simple 2 columns sets, however they have different time stamps.  Your optional assignment is to time-base sequence the data and aggregate based on hour (example of what this looks like, follows).  Note for multiple recordings within an hour, take the mean.  Then to determine if the data is stationary and can it be forecast.  If so, provide a week forward forecast and present results via Rpubs and .rmd and the forecast in an Excel readable file. 



```{r}
# read the data
water1 <- readxl::read_excel("C:/Users/rites/Documents/GitHub/Data624-Project1/Waterflow_Pipe1.xlsx")
water2 <- readxl::read_excel("C:/Users/rites/Documents/GitHub/Data624-Project1/Waterflow_Pipe2.xlsx")

```

#### In order to use the two series together, the readings for pipeline 1 must be converted to hourly:

```{r}
water1 <- water1 %>% 
# separate date & hour components of readings
mutate(Date = date(DateTime),
# convert hour to hour-ending to match pipeline 2
Hour = hour(DateTime) + 1) %>% 
# get average reading for each date & hour
group_by(Date, Hour) %>% 
summarize(WaterFlow = mean(WaterFlow)) %>% 
# convert back to DateTime and drop separate date/hour columns
ungroup() %>%
mutate(DateTime = ymd_h(paste(Date, Hour))) %>% 
select(DateTime, WaterFlow)
```

#### The two datasets are joined and a total wateflow is created, then converted to a timeseries:

```{r}
# create df with both observations for each hour
water_df <- full_join(water1, water2, by = "DateTime", suffix = c("_1", "_2")) %>% 
# convert missing pipeline 1 readings to zero
mutate(WaterFlow_1 = ifelse(is.na(WaterFlow_1), 0, WaterFlow_1)) %>% 
# get total waterflow by hour
mutate(WaterFlow = WaterFlow_1 + WaterFlow_2) %>% 
# drop individual numbers
select(DateTime, WaterFlow)
# create hourly timeseries object
water_ts <- ts(water_df$WaterFlow, frequency = 24)
```

The timeseries is plotted to inspect its features:

```{r}
autoplot(water_ts) +
labs(title = "Hourly water flow through two pipelines", x = "Day", y = "Total waterflow")
```

####There is variability across the whole range, with an initial downward trend before day 10 followed by a roughly flat period through the end of the time window.

#### Fitting

####The variance is constant but we will perform .Due to the apparent non-stationarity, a lag-1 difference is taken:

```{r c-trans}
# Box-cox paramter & transform
water_lambda <- BoxCox.lambda(water_ts)
water_trans <- BoxCox(water_ts, water_lambda)
# plot differenced transformed series
ggtsdisplay(diff(water_trans), points = FALSE,
main = "Differenced Box-Cox transformed water flow")
```

####The timeseries is stationary, but shows significant spikes in the ACF and PACF at $k = 1$, strongly suggesting non-seasonal AR(1) and MA(1) components.  There is no seasonal behavior.  Thus, an ARIMA(1,1,1) model is used:

```{r}
water_fit <- Arima(water_ts, order = c(1, 1, 1), lambda = water_lambda)
```

```{r}
Box.test(resid(water_fit), type = "L")
ggtsdisplay(resid(water_fit), points = FALSE, plot.type = "histogram",
main = "Residuals for ARIMA(1,1,1) of water flow")
```

#### Ljung-Box test returns a value of almost 1; the residuals appear to be roughly normally distributed around 0 without significant autocorrelation.  The model is acceptable and will be used for forecasting.

#### Forecasting

####Using the ARIMA(1,1,1) model, one week (168 hours) is forecast, and the forecast plotted:

```{r}
water_forecast <- forecast(water_fit, 168, level = 95)
autoplot(water_forecast) + 
labs(title = "Forecasted water flow",
x = "Day", y = "Total flow") +
theme(legend.position = "none")
```

#### There is lack of seasonality so a single value is forecasted.  The forecast values are output to a .csv file:

```{r}
data_frame(DateTime = max(water_df$DateTime) + hours(1:168),
WaterFlow = water_forecast$mean) %>% 
write_csv("C:/Users/rites/Documents/GitHub/Data624-Project1/project1_water.csv")
```



